---
title: "A Study of Suicide Rates and Economic Indicators"
author: "Omer Cagri Aydin"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    toc: yes
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stargazer)
library(dplyr)
library(ggplot2)
library(corrplot)
```

# Introduction

A sensitive topic such as suicide of course has many levels and factors to it that must be individually studied for each different case. However, the purpose of this research is to find out if there exists any general correlation between economic indicators and suicide rates. I believe that there might be a correlation between these variables and determining a trend could help suicide prevention groups to target specific groups of people that are in increased risks of self-harm due to the surrounding economic factors. 

The dependent variable in this data set will be suicides.100k.pop (suicide rates per 100.000 people) while my independent variables will be HDI, GDP per capita, and generation. I will study the correlations between my independent variables and suicides.100k.pop. Moreover, I will construct graphs in order to visualize the realtionship between the variables. Finally, I will conduct a regression analysis of relevant variables, conduct a T test, and a F test in order to study the significance. I will also figure out a way to use generation variable in my regression. To do so, I must manipulate this data, which is categorical, into being numerical. 

I believe that there might be a relationship between these valuables due to the way they effect the lifestyle. However, I would not be surprised to find a positive relationship (for example as GDP per capita increases so does suicides per 100k) because of the so called American Paradox. 

Null Hypothesis: H0: u=0
Alternative Hypothesis: H1: u≠0

I will use the following packages for the given reasons

dplyr: rename variables

stargazer:create tables

corrplot: create muti-correlations

ggplot2: plot graphs

# The data set 

I obtained the dataset that I will be using from kaggle.com and the exact source is stated at the bottom of the page. I selected this data set because I am interested in finding out if there exists a relationship between suicide rates and HDI, GDP per capita, and generations. This data is adequate to answer the reseach question because it has a large number of observations and incorporates many countries throughout a large number of years. However, the data first needs some manipulation before it can be used.


```{r }
#setting working directory
setwd("~/Desktop/OCA61/University/Junior/Spring 2019/S19_Eco320L/Final Project")
#importing the data
master <- read.csv("~/Desktop/OCA61/University/Junior/Spring 2019/S19_Eco320L/Final Project/master.csv")
#data manipulation
#selecting only the columns that will be used from the data and creating a new dataset named master1
master1 <- master %>%  select(suicides.100k.pop, HDI.for.year, gdp_per_capita...., generation) 

#I realized that there were a lot of "0" values in the data which corrupted it. This was something I have not realized in my preliminary submission. So I set all of these "0" values to NA
master1[master1==0] <- NA

#Creating a new dataset named master2 in which the NA variables are omitted
master2 <- na.omit(master1)

#This code converts categorical value of generation into numerical. After this I am creating the dataset that I will be using throughout, 
master2p1 <- master2 %>% mutate_if(is.factor, as.numeric)

#Renaming the variables for convinence using dplyr
name<-c(suicides100k ="suicides.100k.pop",hdi="HDI.for.year", gdp.per.capita= "gdp_per_capita....", generation="generation")
master2p1<-rename(master2p1,!!!name)

#summary stats
summary(master2p1)
str(master2p1)

# using stargazer to do a table of the summary statistics of the dataset
stargazer(master2p1,type="text",title="Descriptive Statistics",out="table.txt") 
```

The generations that correspond with the numeric values are:
  
1=Boomers

2=G.I Generation

3=Generation X

4=Generation Z

5=Millenials

6=Silent

# Graphs and Descriptive Statistics 


```{r}
#Creating a Scatterplot for suicides100k and hdi
ggplot(data=master2p1, aes(x=hdi, y=suicides100k)) + geom_point(alpha=0.5, color="red")+
  geom_smooth(method='lm') + ggtitle("Correlation between HDI and Suicide Rates per 100k (linear)") + 
     theme(plot.title = element_text(face="bold"))
```

The scatter plot above shows the correaltion between HDI and the suicide rates per 100k. There seems to be a weak upward trending relation between Suicide Rates and HDI. The data may have no significance. A second scatterplot must be constructed with a non-linear trend line to see in more detail.

```{r}
#Recreating the scatterplot above, however, this time the trend line is nonlinear
ggplot(data=master2p1, aes(x=hdi, y=suicides100k)) + geom_point(alpha=0.5, color="turquoise")+
  geom_smooth() + ggtitle("Correlation between HDI and Suicide Rates per 100k (non-linear)") + 
     theme(plot.title = element_text(face="bold"))
```

There seems to be two spikes on this trend with spikes on 0.7 and 0.8 HDI. However, just by looking at the plot, there still seems to  be no significant relation between HDI and suicide rates. 

```{r}
#Creating a scatterplot with GDP per capita and sucides per 100k
ggplot(data=master2p1, aes(x=gdp.per.capita, y=suicides100k)) + geom_point(alpha=0.7, color="purple")+
  geom_smooth(method='lm', color="yellow") + ggtitle("Correlation between GDP per capita and Suicide Rates per 100k") + 
     theme(plot.title = element_text(face="bold"))
```

There seens to exist a weak downward trend on the line that probably will not be significant. This plot may be tricky. There seems to be a lot more suicide values on the lower side of the GDP per capita. In reality, although there is more data on the left side of the graph, a lot of them come from low suices per 100k values. Because of this, the trend line does not increase a significant decrease in suicide rates.

```{r}
#Creating a histogram for showing the counts of suicide cases for each generation
ggplot(data=master2p1) +
  geom_histogram(mapping=aes(generation), bins = 30, color="black", fill = "red") + ggtitle("Number of Suicides by Generation") + 
     theme(plot.title = element_text(face="bold"))
```
This histogram shows the number of suicides by generation. Looking at the data, generations 2 and 4, G.I. Generation and Generation Z respectfully, seem to have the least count of recorded suicides. 

```{r}
#Constructing a multi-correlation
cor1<-cor(master2p1)
corrplot(cor1,method="number")
```


The multiple correlations above shows that suicides per 100k has no correlation with any of the independent variables. The only correlationthere exits is between hdi and gdp per capita, which is not significant in this study.

# Regression Analysis 

The following models will be created and tested:
$$ y = \beta_0+\beta_1*hdi+u$$
$$ y = \beta_0+\beta_1*hdi + \beta_2*gdp.per.capita+u$$
$$ y = \beta_0+\beta_1*hdi + \beta_2*gdp.per.capita+\beta_3*generation+u$$

```{r}
#Creating linear models
model1<-lm(suicides100k~hdi,data=master2p1)
model2<-lm(suicides100k~hdi+gdp.per.capita,data=master2p1)
model3<-lm(suicides100k~hdi+gdp.per.capita+generation,data=master2p1)

#Estimating the models
stargazer(model1,model2,model3,type="text",title="Models for estimation of Suicide Rates",out="model.txt",dep.var.labels ="Suicide Rates per 100k (suicides100k)",align=TRUE)
```

This estimation generates the following models:


$$y = 4.952+ 11.508*hdi + u$$
$$y = -9.502+33.312*hdi + -0.0001*gdp.per.capita+u$$
$$y = -8.509+34.125*hdi + -0.0001*gdp.per.capita+-0.431*generation+u$$

Looking at these models it could be said that the most accurate model is model2 because model1 has too few variables and the authencity of generation variable in model3 could be disputed. Looking at these modesl, we see that a small incease in hdi can cause to a positive effect in suicide rates, that is as hdi increases so does suicde rates per 100k. Moreover, we see a negative relation with gdp.per.capita. That means that the more the gdp per capita the less the suicide rates. The effect of this could seem small due to the small integer in the front. However, gdp per capita in this data measures as high as 126,352. This means that for certain values above a threshold, the gdp per capita definetely effects the outcome. model3 is no good because I was not able to reflect the generation variable effectively into the equation. The values of this variable hold no significance.

# Inference

```{r}
#running the F-test on all the models
anova(model1,model2,model3)
```

Looking at the results of the F statistic, it could be said that none of the models are able to reject the null hypothesis. model2's F statistic is 65.172 which is significantly larger than 0.05. We fail to reject the null hypothesis. Therefore, it could be said that there is no clear trend between suicide rates per 100k and hdi, gdp per capita, and or generation that the individual belongs to. 


#Sources
https://www.kaggle.com/russellyates88/suicide-rates-overview-1985-to-2016
